beautier: BEAUti for R

### 2018-01-18 article.tex
EOD

@@

\begin{abstract}
  \textbf{1. }
  Here, we present a package, \verb;beautier;, 'BEAUti for R', for the R programming language. \\
  \textbf{2. }
    \verb;beautier; allows for scripted use of the BEAST2 phylogenetics tool, 
    by creating BEAST2 input files from an R function call. \\
  \textbf{3. }
    We describe \verb;beautier; usage, the novel functionality it provides
    compared to BEAUti, and give some minimal examples. \\
  \textbf{4. }
    As \verb;beautier; is free, libre, open-source and designed to be extended, 
    We conclude by describing the current development of the package \\
\end{abstract}

% Key-words: computational biology, evolution, phylogeny, BEAST2, R

@@

BEAST2 \cite{bouckaert2014beast} is a Bayesian phylogenetics tool.
BEAST2 creates a posterior of jointly-estimated phylonies and model
parameters, from a DNA, RNA or amino acid alignment. BEAST2 is a 
console application, that needs a configuration file containing alignments and model parameters.

BEAST2 is bundled with the BEAUti \cite{drummond2012bayesian}. 
BEAUti is a program to create a BEAST2 XML configuration file, 
using a user-friendly graphical user interface, with helpful and reasonable 
default settings. BEAUti replaces the manual editing of the BEAST2 XML
files.

BEAUti cannot be called from a command-line script, which is not a problem in all cases.
For example, the BEAST book \cite{drummond2015bayesian} encourages to first infer a posterior
from simpler models first, then exploring if adding complexity changes the inferred results.
This can easily be done manually using BEAUti. 
For bigger theoretical explorations (for example, using thousands of simulated alignments), this approach
becomes inviable.

Here, we present \verb;beautier;, 'BEAUti for R', which 
creates BEAST2 configuration files from an R function call. 
The interface of \verb;beautier; mimics BEAUti and is easy to use. This
familiar interface helps both beginner and experienced BEAST2 users 
to create configuration files from shell scripts.

@@

  \caption{
    Workflow. From an alignment and priors, one creates a BEAST2 XML input file. This
    can be done using beautier or BEAUti. The created configuration file is run by lumier or BEAST2
    to create a posterior of phylogenies and model parameter estimates.
  }
  \label{fig:workflow}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\verb;beautier;'s main function is \verb;create_beast2_input_file;, which creates
an BEAST2 input file. \verb;create_beast2_input_file; needs at least the name of a FASTA file containing a DNA alignment
and a name for the to-be-created output file. This interface follows BEAUti's default settings.
Per alignment, a site model, clock model and tree priors can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, clock model and tree prior.

@@

In total, \verb;beautier; has 59 exported functions to create
a BEAST2 configuration file. \verb;beautier; is an alternative for a majority of BEAUti use cases.
\verb;beautier; does not support the full functionality of BEAUti. Considering
BEAUti's flexibility and number of plugins, this would be a Herculean effort.
To compensate for this, an extensible software architecture is used.
\verb;beautier;s future extensions can be found on its GitHub.

BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. BEAUti does not allow for fixing
a phylogeny's crown age. Before \verb;beautier;, one needs to manually edit the BEAST2 
XML configuration file, which is prone to errors. \verb;beautier;, allows easy
fixing of phylogenies' crown ages. For theoretical work, a fixed and known crown age can result
in a cleaner analysis.

\verb;beastier; and \verb;lumier; are related packages, used in testing. 
\verb;lumier; calls BEAST2 from within R. \verb;lumier; is used to confirm that the XML files 
created by \verb;beautier; are valid. Additionally, \verb;lumier; is used to run BEAST2
to create posteriors. Using \verb;beastier;, these posteriors are checked to have an
estimated or fixed crown age.

@@

In R, a package's function need to be loaded in the global namespace first:

@@

BEAUti, and likewise \verb;beautier;, need at least a FASTA filename
and an XML output filename. In BEAUti, this is achieved by loading a FASTA file (resulting
in figure \ref{fig:simplest_beauti_usage}), then saving an output file using a common
save file dialog. In \verb;beautier;, the same is achieved by listing \ref{lst:simplest_example}:

@@

\verb;beautier; creates site models, clock models and tree priors with the same default distributions as BEAUti.
For example, a Yule tree prior assumes that birth rate likelihoods follow a uniform distribution, from minus infinity
to infinity. This assumption entails that negative and positive birth rates are just as likely, where a negative birth rate is biologically impossible. 
One may prefer to have an exponential distribution instead, as this would state that birth rates are always positive, and
higher values are less likely than lower values. To do so \verb;beautier; is shown by listing \ref{lst:diff_distr}:

@@

Novel about \verb;beautier; is that it allows for specifying a fixed crown age. 
By default, a phylogeny's crown age is jointly-estimated with the other parameters.
Setting a fixed crown age is not yet possible in BEAUti directly, but it is documented how to 
manually edit the XML file to allow for a fixed crown age. 
Listing \ref{lst:fixed_crown_age} shows how to specify a fixed crown age with \verb;beautier;:

@@

\verb;beautier;'s development takes place on GitHub \cite{github}, 
which is a good practice for computational scientists \cite{perez2016ten} 
and improves transparency \cite{gorgolewski2016practical}.

>>>>

% Thanks Raphael Scherrer
\usepackage[switch]{lineno}

@@

% TODO: RS: Rename to 'Summary 
\begin{abstract}
  \textbf{1. }
    Here, we present a free, libre and open-source package, \verb;beautier;, 
    'BEAUti for R', for the R programming language. \\
  \textbf{2. }
    \verb;beautier; allows for scripted use of the BEAST2 phylogenetics tool, 
    by creating BEAST2 input files from an R function call. \\
  \textbf{3. }
    We describe \verb;beautier;'s usage, the novel functionality it provides
    compared to BEAUti, and give some minimal examples. \\
  \textbf{4. }
    As \verb;beautier; is designed to be of high quality and extendable, 
    we conclude by describing the development of the package \\
\end{abstract}

% TODO: RS: Get these keywords visualized
% Key-words: computational biology, evolution, phylogeny, BEAST2, R

@@

BEAST2 \cite{bouckaert2014beast} is a Bayesian phylogenetics tool.
BEAST2 creates a posterior of jointly-estimated phylogenies and model
parameters, from a DNA, RNA or amino acid alignment (see figure \ref{fig:workflow} for an overview
of the workflow). BEAST2 is a 
console application, that needs a configuration file containing alignments and model parameters.

BEAST2 is bundled with BEAUti \cite{drummond2012bayesian}. 
BEAUti is a program to create a BEAST2 XML configuration file, 
using a user-friendly graphical user interface, with helpful and reasonable 
default settings. BEAUti replaces the manual editing of the BEAST2 XML
files.

BEAUti cannot be called from a command-line script, which is not a problem in all cases.
For example, the BEAST book \cite{drummond2015bayesian} encourages a stepwise approach: 
(1) infer a posterior from the simplest models possible, (2) while a posterior with a more complex model gives
a different posterior, increase complexity. This can be done manually using BEAUti and is the common workflow
when using a few alignments and doing a superficial search through model space.

Here, we present \verb;beautier;, 'BEAUti for R', which 
creates BEAST2 configuration files from an R function call.
The interface of \verb;beautier; mimics BEAUti and is easy to use. This
familiar interface helps both beginner and experienced BEAST2 users 
to create configuration files from shell scripts.
Using \verb;beautier; will save you time and thousands of mouse clicks.
\verb;beautier; brings deep exploration of model space within reach,
as you can loop through multiple (simulated) alignments, 
nucleotide substitution models, clock models and tree priors.

@@

  \caption{
    Workflow. From an alignment and priors (depicted as a thought cloud), one creates a BEAST2 XML input file. This
    can be done using beautier or BEAUti. The created configuration file is run by lumier or BEAST2
    to create a posterior of phylogenies and model parameter estimates.
  }
  \label{fig:workflow}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\verb;beautier;'s main function is \verb;create_beast2_input_file;, which creates
a BEAST2 input file. \verb;create_beast2_input_file; needs at least the name of a FASTA file containing a DNA alignment
and a name for the to-be-created output file. This interface follows BEAUti's default settings.
Per alignment, a site model, clock model and tree prior can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, clock model and tree prior.

@@

In total, \verb;beautier; has 59 exported functions to create
a BEAST2 configuration file. \verb;beautier; is an alternative for a majority of BEAUti use cases.
\verb;beautier; does not support the full functionality of BEAUti. 
Because of its high number of plugins, an extensible software architecture is used.
A roadmap of \verb;beautier;'s future extensions can be found on its GitHub.

BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. BEAUti does not allow for fixing
a phylogeny's crown age. Without \verb;beautier;, one needs to manually edit the BEAST2 
XML configuration file, which is prone to errors. \verb;beautier;, allows easy
fixing of phylogenies' crown ages. For theoretical work, a fixed and known crown age can result
in a cleaner analysis.

\verb;beastier; and \verb;lumier; are related packages, which are used to confirm the correct
working of verb;beautier;. \verb;lumier; calls BEAST2 from within R and is used to confirm that the XML files 
created by \verb;beautier; are valid. Additionally, \verb;lumier; is used to run BEAST2
to create posteriors. Using \verb;beastier;, these posteriors are checked to have an
estimated or fixed crown age.

@@

In R, a package's function needs to be loaded in the global namespace first:

@@

BEAUti, and likewise \verb;beautier;, needs at least a FASTA filename
and an XML output filename. In BEAUti, this is achieved by loading a FASTA file (resulting
in figure \ref{fig:simplest_beauti_usage}), then saving an output file using a common
save file dialog. In \verb;beautier;, the same is achieved by listing \ref{lst:simplest_example}:

@@

\verb;beautier; creates site models, clock models and tree priors with the same default distributions as BEAUti.
For example, a Yule tree prior assumes that birth rate likelihoods follow a uniform distribution, from minus infinity
to infinity. This assumption entails that negative and positive birth rates are just as likely, where a negative birth rate is biologically impossible. 
One may prefer to have an exponential distribution instead, as this would state that birth rates are always positive, and
higher values are less likely to occur than lower values. 
Listing \ref{lst:diff_distr} shows how to do so in \verb;beautier;:

@@

Novel functionality that \verb;beautier; offers, is that it allows for specifying a fixed crown age. 
By default, a phylogeny's crown age is jointly-estimated with the other parameters.
Setting a fixed crown age is not yet possible in BEAUti directly, but it is documented how to 
manually edit the XML file to allow for a fixed crown age. 
Listing \ref{lst:fixed_crown_age} shows how to specify a fixed crown age with \verb;beautier;:

@@

\verb;beautier;'s development takes place on GitHub \cite{github}, 
facilitating collaboration \cite{perez2016ten} 
and improving transparency \cite{gorgolewski2016practical}.
